Marlowe: implementing and analysing financial contracts on blockchain

### 2019-12-11 papers/marlowe2019/paper.tex

\texttt{Value}s include some quantities that change with time, including the current slot number, the current balance of an account (in Lovelace), and any choices that have already been made. % ; we call these volatile values. - PLS: this definition is no longer used
Values can also be combined using addition, subtraction and negation.

@@

\texttt{Pay}: A payment contract \texttt{Pay a p v cont} will make a payment of value \texttt{v} from the account \texttt{a} to a payee \texttt{p}, which will be one of the contract participants or another account in the contract. Warnings will be generated if the value \texttt{v} is not positive, or if there is not enough in the account to make the payment in full. In the first case, nothing will be transferred; in the later case, a partial payment (of all the money available) is made. The contract will continue as specified by \texttt{cont}.

@@

Marlowe is specified by an executable semantics written in Haskell, but to make it usable in practice to execute financial contracts, it needs to be implemented on a blockchain. In this section, we explain how Marlowe is executed on the Cardano blockchain using an interpreter \footnote{The implementation is available at \url{https://github.com/input-output-hk/plutus/blob/0ca9af4f6614d591de7ebbe4dd759ce122d74efd/marlowe/src/Language/Marlowe/Semantics.hs}.} written in the Plutus programming language.

@@

There are several ways to implement Marlowe contracts on top of Plutus.
We could write a Marlowe to Plutus compiler that would convert
each Marlowe contract into a specific Plutus script.
Instead, we chose to implement an interpreter of Marlowe contracts as a single Plutus script.
This approach has a number of advantages:
\begin{itemize}
\item It is simple: having a single Plutus script that implements all Marlowe contracts
makes it easier to implement, review, and test what we have done.
\item Implementation is close to the semantics of Marlowe, as sketched above and in more detail in~\cite{isola-marlowe},
which makes it easier to validate.
\item The same implementation can be used for both on- and off-chain (wallet) execution of Marlowe code.
\item It facilitates client-side contract evaluation, where we reuse the same code
to do contract execution emulation in an IDE, and compile it to WASM/JavaScript on the client side,
e.g.\ in the Marlowe Playground.
\item Having a single interpreter for all (or a particular group of) Marlowe contracts allows us
  to monitor the blockchain for these contracts, if required.
\item Finally, Cardano nodes could potentially use an optimised interpreter (e.g:~native) just for Marlowe contracts, which would save processing time.
\end{itemize}

@@

The Marlowe interpreter first validates the current contract state: i.e.\
we check that the contract locks at least what it should, according
to contract balances (the \texttt{accounts} field in \texttt{State}),
and that all balances are strictly positive.
\footnote{
Using the Isabelle proof assistant, we have formally verified that given a state with positive balances,
it is impossible for any possible contract and inputs to result in non-positive balances.
This is described in more detail in Section \ref{sec:static}.}

@@

Our implementation relies on the Haskell library SBV, which in turn relies on existing SMT solvers to check satisfiability of properties or finding counter-examples.

@@

SBV provides a monad called \texttt{SBV}, a function can use parameters wrapped in this monad to represent symbolic values. Functions that take symbolic values can be used as properties and passed to the solver; the solver will then try to replace the symbolic values with concrete values that satisfy or falsify the property.

@@

Isabelle automatically proves termination for most function. This is not the case for \texttt{reductionLoop}. This function repeatedly calls \texttt{reduceContractStep} until it returns \texttt{NotReduced}, so proving overall termination requires a proof that \texttt{reduceContractStep} will eventually do that.

We proved that by establishing a measure for the size of a pair of \texttt{Contract} and \texttt{State}:

\begin{verbatim}
fun evalBound :: "State ⇒ Contract ⇒ nat" where
"evalBound sta cont = length (accounts sta) + 2 * (size cont)"
\end{verbatim}
\noindent
where \texttt{size} is a measure already generated automatically by Isabelle for custom data types.

@@

\subsection{Valid state and positive account preservation}

There are some values for \texttt{State} that are allowed by its type but make no sense, especially in the case of Isabelle semantics where we use lists instead of maps and sets:

\begin{enumerate}
    \item The lists represent maps, so they should have no repeated keys.
    \item We want two maps that are equal to be represented the same, so we force keys to be in ascending order.
    \item For the map of accounts, we only want to record those that have positive amounts.
\end{enumerate}
\noindent
We call a value for \texttt{State} \emph{valid} if the first two properties are true. And we say it has \emph{positive accounts} if the third property is true.

@@

A contract is \emph{quiescent} if and only if the root construct \texttt{When}, or if the contract is \texttt{Close} and all accounts are empty.

We have proved that, if an input \texttt{State} is valid and all accounts are positive, then the result from processing any transaction will always be quiescent.

@@

\subsubsection{Money preservation}
Money is not created or destroyed by the semantics. More specifically, the money that comes in plust the money in the contract before the transaction must be equal to the money that comes out plus the contract after the transaction, except in the case of an error, when the contract stays unaltered.

@@

We proved that this conservative upper bound is general enough for every contract, by showing that, if the contract is not closed and empty, then an empty transaction sent after \texttt{maxTime} will close the contract and leave the accounts empty.

Because of this property and \emph{Money preservation}, we can conclude that money will not be locked in the contract forever.

@@

This property implies that any trace that is longer than this is guaranteed to produce at least one error. Because transactions that produce an error do not alter the state of the contract, such a list of transactions (a trace) will be equivalent to a list of transactions that does not have the erroneous transaction.

Thus, we do not lose generality by only exploring traces with length of at most \texttt{maxTransactionsInitialState}, because any longer traces produce the same effect.

@@

On the simulation pane, ``pure'' Marlowe contacts can be developed: that is, contracts not embedded in a host language. The reason for doing this is two-fold:
\begin{itemize}
    \item There is a shallower learning curve for users who are new to Haskell or programming languages in general. The Marlowe constructs are quite simple, and there is no need, at least initially, to learn about Haskell syntax or even variables, functions etc.
    \item As we step through the execution of a contract in a simulation, the contract is reduced; it is very useful to be able to view, or even edit, the reduced contract during this execution.
\end{itemize}
As contracts become larger it makes sense to use another editor in the \emph{Haskell} pane. Here contracts can be written using facilities from \emph{Haskell} to abbreviate and make more readable the description of the contracts. These contracts can then be transferred as a pure Marlowe data structure into the simulation pane.

Contracts can also be written using Google's \emph{Blockly} visual programming language, as was earlier described in Meadow~\cite{isola-marlowe}. Blockly gives an easy way to introduce the concepts of Marlowe to users who have no programming knowledge, and in particular the editor gives users a set of options for each construct as the contract is built. Once a contract has been constructed in Blockly it is possible to transfer that contract to the simulation pane.%SJT Changed from "Marlowe editor".
It is also possible to transfer a Marlowe contract to Blockly for further editing.

\begin{wrapfigure}[9]{l}{0.25\textwidth}
    \vspace*{-0.3in}
    \includegraphics[width=0.25\textwidth]{hole_options.png}
\end{wrapfigure}

The Marlowe editor in the unstable version of the playground has an additional feature to aid writing contracts called holes. If we enter the contract \lstinline{?mycontract}
%SJT: enter it where?
we will be presented with a dropdown list of values that could be used.

@@

\begin{wrapfigure}[6]{l}{0.25\textwidth}
    \vspace*{-0in}
    \includegraphics[scale=0.2]{hole_options_2.png}
\end{wrapfigure}
A complete contract can be written in this guided way with the user needing only to fill in strings and numbers by hand. This approach to writing holes in your code and ``asking'' the compiler what you could put in there is easy to implement in a DSL because there are very limited options, however is is also becoming popular with more complex languages such as Haskell and Idris.

@@

\begin{wrapfigure}[6]{l}{0.5\textwidth}
    \vspace*{-0.1in}
    \includegraphics[scale=0.15]{input_composer.png}
\end{wrapfigure}

@@

At any point in the simulation the user can undo any steps made: in this particular case, they can undo the application of the transaction, and iteratively undo more steps. At any point they can also reset the contract to its initial state. This enables a user to apply transactions, see what the effect is, step back and apply a transaction with different actions and see how these differences affect the end result. They can also change the reduced contract to investigate variants of the original.

@@

Defining the language by means of an executable reference semantics means that we can, as well as directly executing this semantics, generate an on-chain interpreter for it and simulate it in browser using the Haskell-like languages Plutus and PureScript. This is made particularly straightforward by working in the subset of Haskell that directly corresponds to what translates directly into these languages.

>>>>

\texttt{Value}s include some quantities that change with time, like the current slot number, the current balance of an account, and any choices that have already been made. % ; we call these volatile values. - PLS: this definition is no longer used
Values can be combined using addition, subtraction and negation.

@@

\texttt{Pay}: A payment contract \texttt{Pay a p v cont} will make a payment of value \texttt{v} from the account \texttt{a} to a payee \texttt{p}, which will be one of the contract participants or another account in the contract. Warnings will be generated if the value \texttt{v} is not positive, or if there is not enough in the account to make the payment in full. In the first case, nothing will be transferred; in the later case, a partial payment (of all the money available) is made. The contract will continue as \texttt{cont}.

@@

Marlowe is specified by an executable semantics written in Haskell, but to make it usable in practice with financial contracts, it needs to be implemented on a blockchain. In this section, we explain how Marlowe is executed on the Cardano blockchain using an interpreter \footnote{The implementation is available at \url{https://github.com/input-output-hk/plutus/blob/0ca9af4f6614d591de7ebbe4dd759ce122d74efd/marlowe/src/Language/Marlowe/Semantics.hs}.} written in the Plutus programming language.

@@

There are several ways to implement Marlowe contracts on top of Plutus.
We could write a Marlowe to Plutus compiler that would convert
each Marlowe contract into a specific Plutus script.
Instead, we chose to implement a Marlowe interpreter as a single Plutus script.
This approach has a number of advantages:
\begin{itemize}
\item It is simple: having a single Plutus script that implements all Marlowe contracts
makes it easier to implement, review, and test what we have done.
\item Implementation is close to the semantics of Marlowe, as sketched above and in more detail in~\cite{isola-marlowe},
which makes it easier to validate.
\item The same implementation can be used for both on- and off-chain (wallet) execution of Marlowe code.
\item It facilitates client-side contract evaluation, where we reuse the same code
to do contract execution emulation in an IDE, and compile it to WASM/JavaScript on the client side,
e.g.\ in the Marlowe Playground.
\item Having a single interpreter for all (or a particular group of) Marlowe contracts allows us
  to monitor the blockchain for these contracts, if required.
\item Finally, Cardano nodes could potentially use an optimised interpreter (e.g:~native) just for Marlowe contracts, which would save processing time.
\end{itemize}

@@

The Marlowe interpreter first validates the current contract state: i.e.\
we check that the contract locks at least as much as specified by the
contract balances (the \texttt{accounts} field in \texttt{State}),
and that balances are strictly positive.
\footnote{
Using the Isabelle proof assistant, we have formally verified that given a state with positive balances,
it is impossible for any possible contract and inputs to result in non-positive balances.
This is described in more detail in Section \ref{subsec:valid_state}.}

@@

Our implementation relies on the Haskell library SBV, which in turn relies on existing SMT solvers to check satisfiability of properties.

@@

SBV provides a monad called \texttt{SBV}, a function can use parameters wrapped in this monad to represent symbolic values. Functions that take symbolic values can be used as properties and passed to the solver; which will replace the symbolic values with concrete values that satisfy or falsify the property.

@@

Isabelle automatically proves termination for most function. This is not the case for \texttt{reductionLoop}. This function repeatedly calls \texttt{reduceContractStep} until it returns \texttt{NotReduced}, so proving overall termination requires a proof that \texttt{reduceContractStep} will eventually do that. In order to prove this, we defined a measure for the size of a pair of \texttt{Contract} and \texttt{State}:

\begin{verbatim}
fun evalBound :: "State ⇒ Contract ⇒ nat" where
"evalBound sta cont = length (accounts sta) + 2 * (size cont)"
\end{verbatim}
\noindent
where \texttt{size} is a measure already generated automatically by Isabelle.

@@

\subsection{Valid state and positive account preservation\label{subsec:valid_state}}

There are some values for \texttt{State} that are allowed by its type but make no sense, especially in the case of Isabelle semantics where we use lists instead of maps:

\begin{enumerate}
    \item The lists represent maps, so they should have no repeated keys.
    \item We want two maps that are equal to be represented the same, so we force keys to be in ascending order.
    \item We only want to record those accounts that have positive amount.
\end{enumerate}
\noindent
We call a value for \texttt{State} \emph{valid} if the first two properties are true. And we say it has \emph{positive accounts} if the third property is true.

@@

A contract is \emph{quiescent} if and only if the root construct \texttt{When}, or if the contract is \texttt{Close} and all accounts are empty. We have proved that, if an input \texttt{State} is valid and accounts are positive, then the output will be quiescent.

@@

\subsubsection{Money preservation}
Money is not created or destroyed by the semantics. More specifically, the money that comes in plus the money in the contract before the transaction must be equal to the money that comes out plus the contract after the transaction, except in the case of an error.

@@

We proved that this conservative upper bound is general enough for every contract, by showing that, if the contract is not closed and empty, then an empty transaction sent after \texttt{maxTime} will close the contract and empty the accounts.

@@

This property implies that any trace that is longer than this is guaranteed to produce at least one error. Because transactions that produce an error do not alter the state of the contract, such a list of transactions (a trace) will be equivalent to a list of transactions that does not have the erroneous transaction.
Thus, we do not lose generality by only exploring shorter traces.

@@

On the simulation pane, ``pure'' Marlowe contacts can be developed: that is, contracts not embedded. Two reasons for doing this are:
\begin{itemize}
    \item There is a shallower learning curve for users who are new to Haskell or programming languages in general. The Marlowe constructs are quite simple, and there is no need, at least initially, to learn about Haskell syntax or even variables, functions etc.
    \item As we step through the execution of a contract in a simulation, the contract is reduced; it is very useful to be able to view, or even edit, the reduced contract during this execution.
\end{itemize}
As contracts become larger it makes sense to use another editor in the \emph{Haskell} pane. Here contracts can be written using facilities from \emph{Haskell} to abbreviate and make more readable the description of the contracts. These contracts can then be transferred as a pure Marlowe data structure into the simulation pane.

Contracts can also be written using Google's \emph{Blockly} visual programming language, as was earlier described in Meadow~\cite{isola-marlowe}. Blockly gives an easy way to introduce the concepts of Marlowe to users who have no programming knowledge, and in particular the editor gives users a set of options for each construct as the contract is built. Once a contract has been constructed in Blockly it is possible to transfer that contract to the simulation pane. It is also possible to transfer a Marlowe contract to Blockly for further editing.

\begin{wrapfigure}[9]{l}{0.25\textwidth}
    \vspace*{-0.28in}
    \includegraphics[width=0.25\textwidth]{hole_options.png}
\end{wrapfigure}

The Marlowe editor in the unstable version of the playground has a feature called \emph{holes} to aid writing contracts. If we enter the contract \lstinline{?mycontract}
%SJT: enter it where?
we will be presented with a dropdown list of values that could be used.

@@

\begin{wrapfigure}[6]{l}{0.25\textwidth}
    \vspace*{-0.00in}
    \includegraphics[scale=0.2]{hole_options_2.png}
\end{wrapfigure}
A complete contract can be written in this guided way with the user needing only to fill in strings and numbers by hand. This approach to writing holes in your code and ``asking'' the compiler what you could put in there is easy to implement in a DSL because there are very limited options, however is is also becoming popular with more complex languages such as Haskell and Idris.

@@

\begin{wrapfigure}[5]{l}{0.5\textwidth}
    \vspace*{-0.2in}
    \includegraphics[scale=0.15]{input_composer.png}
\end{wrapfigure}

@@

At any point in the simulation, the user can undo any steps made: in this particular case, they can undo the application of the transaction, and iteratively undo more steps. At any point, they can also reset the contract to its initial state. This enables users to apply transactions, see their effects, step back, and try different transactions to see the effects of the changes on the result. They can also change the reduced contract to investigate variants of the original.

@@

Defining the language by means of an executable reference semantics means that we can, as well as directly executing this semantics, generate an on-chain interpreter for it and simulate it in browser using the Haskell-like languages Plutus and PureScript. This is particularly straightforward when working with a subset of Haskell that is represented in the same way on these languages.
